\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-09 challenge.tex
sound logical reasoning

Inheritance and family polymorphism are mechanisms for
incrementally modifying self-referential definitions~\cite{cook1990inheritance}:
the implicit self parameterization is how they achieve late binding and
thus offer extensibility.
However, self reference could easily lead to divergence.
Divergence is not a concern for the design of ordinary OO or
functional languages, but it would mean logical inconsistency---and
hence unsoundness!---for a language aimed at logical reasoning.
A family-polymorphism design must tame self reference to guarantee consistency.

>>>>

The language-theoretic essence of late binding is self reference;
inheritance and family polymorphism are mechanisms for
incrementally modifying self-referential definitions~\cite{cook1990inheritance}.
%the implicit self parameterization is how they achieve late binding and
%thus offer extensibility.
However, self reference could easily lead to divergence.
Divergence is not a concern for the design of ordinary OO or
functional languages, but it would mean logical inconsistency---and
hence unsoundness!---for a language aimed at logical reasoning.
A family-polymorphism design must tame self reference to guarantee consistency.

### 2022-10-09 lang-design.tex
sound logical reasoning

Note that definitional equality on \lsti{subst} is available
outside those families that contain "subst".
Within those families, the meaning of \lsti{subst} is late bound.
In contrast, outside those families, \lsti{subst} can be referenced only
by specifying a family that contains it,
such as \lsti{STLC.subst} or \lsti{STLC_fix.subst}.
So as far as Coq's type checker is concerned, \lsti{tm_unit} and
\lsti{STLC_fix.subst tm_unit x t'} are the same thing---Coq equates them
definitionally by unfolding \lsti{STLC_fix.subst} and performing
$\beta$-reduction.

@@

\noindentparagraph{Ensuring consistency.}

Example of a circular proof

>>>>

Note that the definitional equality on \lsti{subst} is available
outside those families that contain "subst".
Within those families, the meaning of \lsti{subst} is late bound.
In contrast, outside those families, \lsti{subst} can be referenced only
by specifying a family that contains it,
such as \lsti{STLC.subst} or \lsti{STLC_fix.subst}.
So as far as Coq's type checker is concerned, \lsti{tm_unit} and
\lsti{STLC_fix.subst tm_unit x t'} are the same thing---Coq equates them
definitionally by unfolding \lsti{STLC_fix.subst} and performing
$\beta$-reduction.

@@

In an OO language, a subclass can override methods of a superclass.
Similar expressiveness is useful for metatheory mechanization, too.
%
For example, in a derived family, the programmer may want to override
the proof of a lemma with a different proof strategy than it was proved
in the base family.
%
Overriding is potentially in conflict with having equalities on late bound names.

Coq distinguishes \emph{opaque} definitions from \emph{transparent} ones.
Our plugin supports the overriding of opaque definitions, which
include most proofs.
It is safe to override opaque definitions, because Coq will never attempt to unfold them.
For transparent definitions, the common case is that the programmer does
not want to override them.
In \cref{fig:stlc-mechanized}, "env", "empty", "subst", and "steps" are
transparent---that is, they are not defined with \lsti{Qed}.
Our plugin treats transparent definitions as non-overridable by default.
Thus, the definitional equalities on "env", "empty", and "steps",
as well as the propositional equality on "subst", are available to the
type checker for type-checking the families.

Our plugin does allow the overriding of transparent definitions that are
explicitly marked as \lsti{Overridable} by the programmer.
Overriding is made safe by requiring that when an overridable field is
overridden, code whose type checking requires unfolding that field
should be overridden too.
We give an example in a supplemental appendix.

\noindentparagraph{Sound logical reasoning.}

In \cref{fig:stlc-mechanized}, just because the "typesafe" theorem in
"STLC" is inherited and reused by "STLC_fix", it does not follow that
in "STLC_fix" the programmer can use "typesafe" to prove "progress".
If the programmer did, then they would be committing the logical fallacy
of circular reasoning: the proof of "progress" would depend on "typesafe",
yet the proof of "typesafe" depends on "progress".
%
Such circularity would easily lead to logical inconsistency:

\begin{centered}
\begin{minipage}{36ex}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
Family A.
FLemma f : False. Proof. Admitted.
FLemma g : False. Proof. apply f. Qed.
End A.
\end{lstlisting}
\end{minipage}
\hspace{3.5em}
\begin{minipage}{36ex}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
Family B extends A.
FLemma f : False. Proof. apply g. Qed.
End A.
[@~@]
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
In the program above, family~"B" extends~"A", overriding lemma~"f" by
proving it using lemma~"g".
Lemma~"g" is in turn inherited from family "A", where it is proven using
a late bound reference to lemma~"f".
Circularity between "f" and "g" allows proving \lsti{False}!

To ensure the soundness of logical reasoning, the type system requires
that in a derived family, the context in which a field is defined be
preserved from the base family.
In "STLC", "progress" is in the context of "typesafe".
Per the requirement, this relationship must be preserved into
"STLC_fix", which prevents the proof of "progress" from depending on
"typesafe" in "STLC_fix".

Note that the requirement still allows a derived family to introduce
new dependencies into the context of a field.
For example, consider an extension of STLC with iso-recursive types
(left column of \cref{fig:stlc-isorec-prod}).
Family "STLC_isorec" introduces "tysubst" into the context of "hasty".
\input{stlc-isorec-prod.tex}

### 2022-10-09 stlc-mechanized.tex
sound logical reasoning

\lstset{
  basicstyle=\fontsize{8}{8.5}\ttfamily,
}

\begin{lstlisting}[numbers=left]
Family STLC.

@@

Field env : Type := id -> option ty.
Field empty : env := \lambda _, None.

@@

Field steps := clos_refl_trans step.

@@

FInduction substlem on hasty[@\hfill\codecomment{Subst.\ lemma}@]
motive \lambda G' t T (_ : hasty G' t T),
  \forall G x t' T', G' = extend G x T' ->
  hasty empty t' T' -> hasty G (subst t x t') T.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End substlem.

FInduction preserve on hasty[@\hfill\codecomment{Preserv.\ theorem}@]
motive \lambda G t T (_ : hasty G t T),
  G = empty -> \forall t', step t t' -> hasty empty t' T.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End preserve.

@@

Field typesafe :[@\hfill\codecomment{Type-safety theorem}@]
  \forall t t' T, steps t t' -> hasty empty t T ->
  value t' \/ \exists t'', step t' t''.
Proof. [@\dadada@] Qed.

End STLC.
\end{lstlisting}
\medskip

\hrule
\hrule
\medskip

\begin{lstlisting}[numbers=left,firstnumber=77]
Family STLC_fix extends STLC.

FInductive tm : Set +=
| tm_fix : id -> tm -> tm.

@@

End STLC_fix.
\end{lstlisting}
\medskip

\hrule
\hrule
\medskip

\begin{lstlisting}[numbers=left,firstnumber=111]
Check STLC_fix.typesafe.
\end{lstlisting}

@@

 Extend FInductive step : term -> term -> Prop 
  := ... (* Need to expand this binary relation *)
(* ... and more extension *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \caption{Example STLC and its extension, in designed syntax}\label{fig:STLC-example2}
\end{figure}

>>>>

\lstset{
  basicstyle=\fontsize{8}{8.5}\ttfamily,
% numbers=left,
}

\colorlet{codecomment-color}{dkgreen}

\begin{lstlisting}
Family STLC.[@\hfill\codecomment{The base STLC}@][@\tanc{STLC-start}@]

@@

FDefinition env : Type := id -> option ty.
FDefinition empty : env := \lambda _, None.

@@

FDefinition steps := clos_refl_trans step.

@@

FInduction substlem on hasty[@\hfill\codecomment{Subst.\ lemma}@]
motive \lambda G' t T (_ : hasty G' t T),
  \forall G x t' T', G' = extend G x T' ->
  hasty empty t' T' -> hasty G (subst t x t') T.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End substlem.[@\tanc{STLC-fstcol-end}\columnbreak@]
FInduction preserve on hasty[@\hfill\codecomment{Preserv.\ theorem}@][@\tanc{STLC-sndcol-start}@]
motive \lambda G t T (_ : hasty G t T),
  G = empty -> \forall t', step t t' -> hasty empty t' T.
Case ht_unit. [@\dadada@] Qed.[@\hfill@]Case ht_var. [@\dadada@] Qed.
Case ht_abs.  [@\dadada@] Qed.[@\hfill@]Case ht_app. [@\dadada@] Qed.
End preserve.

@@

FTheorem typesafe :[@\hfill\codecomment{Type-safety theorem}@]
  \forall t t' T, steps t t' -> hasty empty t T ->
  value t' \/ \exists t'', step t' t''.
Proof. [@\dadada@] Qed.

End STLC.[@\tanc{STLC-end}@]
\end{lstlisting}

\makeline[0pt]{STLC-start}{STLC-fstcol-end}[codecomment-color!50]
\makeline[.5\textwidth+9pt]{STLC-sndcol-start}{STLC-end}[codecomment-color!50]

\vspace{-8pt}

\colorlet{codecomment-color}{orange}

\begin{lstlisting}[firstnumber=77]
Family STLC_fix extends STLC.[@\tanc{STLCfix-start}@]
[@\hfill\codecomment{STLC extended with fixpoints}@]
FInductive tm : Set +=
| tm_fix : id -> tm -> tm.

@@

End STLC_fix.[@\tanc{STLCfix-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCfix-start}{STLCfix-end}[codecomment-color!50]

\vspace{-8pt}

\colorlet{codecomment-color}{black}

\begin{lstlisting}[firstnumber=111]
Check STLC_fix.typesafe.[@\tanc{Check-STLCfix-typesafe}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{Check-STLCfix-typesafe}{Check-STLCfix-typesafe}[codecomment-color!35]

@@

 Extend FInductive step : term -> term -> Prop 
  := ... (* Need to expand this binary relation *)
(* ... and more extension *)
EndFamily.
\end{lstlisting}
\end{minipage}

\caption{Example STLC and its extension, in designed syntax}\label{fig:STLC-example2}
\end{figure}
